\documentclass[handout]{beamer}
%\documentclass[handout]{beamer}

\include{base}

\title{CUCaTS Python Workshop}
\date{Lent 2015}

\begin{document}
	\begin{frame}[fragile]
		\frametitle{Making change}
		\footnotesize
		Cast your mind back to Friday night. The scene is the ADC bar. It's late, you've had a few.
		
		But not enough! So you wander over to the barman and mutter something about a Staropramen. ``2.65 please".
		
		You reach for your wallet, but crap! The ADC doesn't take cards!
		
		You rustle through your pockets and pull out the contents. No notes, but a whole load of coins. 5ps, 10ps, 20ps. You look at it and think to yourself ``I'm... sure this is enough to make 2.65, but the question is... how?"
		
		That is the question we will answer today, and we will answer it with Python!
	\end{frame}

	\begin{frame}[fragile]
		\frametitle{Problem data}
		
		Going back to the first session: we told you that programming was about processing data. In this problem, what data are we going to be dealing with?
		
		\pause
		
		Input:
		\begin{itemize}
			\item \lstinline|amount| \\
			      How much we want to make
			\item \lstinline|coins| \\
			      The coins that we have to make it
		\end{itemize}
		
		\pause
		
		Output:
		\begin{itemize}
			\item \lstinline|ways| \\
			      All the ways of making the amount
				\begin{lstlisting}
				[[100, 5, 20, 20, 20, 20, 20, 20, 20, 20],
				 [200, 50, 10, 5]]
				\end{lstlisting}
		\end{itemize}
	\end{frame}
		
	\begin{frame}[fragile]
		\frametitle{Problem data - Types}
		
		Let's be more specific about our data. Input:
		
		\begin{itemize}
			\item The amount will simply be a number, either integer or float. For simplicity, we'll say its the number of pennies, so an integer.
			\item \lstinline|amount|: integer \\
			\item The coins we have will then just be a list of integers.
			\item \lstinline|coins|: list of integer \\
		\end{itemize}
		
		Output:
		\begin{itemize}
			\item  If we were working out just a \emph{single} way of making the amount, it would be a list of integers, that is, all the coins needed. But we're working out all the possible ways, so our result will be a list of list of integers: each element of the list will itself be another way of making the amount, a list of integers.
			\item \lstinline|ways|: list of list of integers \\
		\end{itemize}
		
	\end{frame}
	
	
	\begin{frame}[fragile]
		\frametitle{}
		
		Writing that up as Python:
		
		\begin{lstlisting}
		def makechange(amount, coins):
		    ways = []
		
		    $\textnormal{[\dots work out all the ways\dots]}$
		
		    return ways
		\end{lstlisting}
		
		Start by filtering out all the coins that are too large to make the amount from our list of \lstinline|coins|. The (an) answer is on the next slide, but try it yourself before looking!
	\end{frame}
	
	\begin{frame}[fragile]
		\frametitle{}
		
		I've used a list comprehension to reproduce the list of coins, but leaving only ones that can fit in \lstinline|amount|.
		
		\begin{lstlisting}[basicstyle=\tt\footnotesize]
		def makechange(amount, coins):
		    ways = []
		
		    #Get rid of the coins that are too large
		    coins = [coin for coin in coins if coin <= amount]
		
		    $\textnormal{[\dots]}$
		
		    return ways
		\end{lstlisting}
	\end{frame}
	
	\begin{frame}[fragile]
		\frametitle{}
		
		Now we'll go through our list of coins in turn\dots
		
		\begin{lstlisting}
		def makechange(amount, coins):
		    ways = []
		
		    #Get rid of the coins that are too large
		    coins = [coin for coin in coins if coin <= amount]
		
		    for coin in coins:
		        $\textnormal{[\dots]}$
		
		    return ways
		\end{lstlisting}
	\end{frame}
	
	\begin{frame}[fragile]
		\frametitle{}
		
		\begin{lstlisting}
		def makechange(amount, coins):
		    #Deal with the trivial case of making zero
		    if amount == 0:
		        return [[]]
		
		    else:
		        ways = []
		
		    #Get rid of the coins that are too large
		    coins = [coin for coin in coins if coin <= amount]
		
		    for coin in coins:
		        ways += also(coin, makechange(amount-coin, coins))
		
		        #Exclude this coin from any more solutions
		        coins = coins[1:]
		
		    return ways
		\end{lstlisting}
	\end{frame}
	
\end{document}